---
title: "Exploratory Analysis"
author: Yuen
date: 10/28/2021
output: 
  github_document
---

##define the theme for all the ggplots!

```{r setup}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.color = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_color.discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r}
#load the weather dataset
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    #round the date to the nearest month with floor_date()
    month = lubridate::floor_date(date,unit = "month")) %>%
  select(name, id, everything())

weather_df

```

## group_by

```{r}
weather_df %>% 
  group_by(name, month) 
  #ungroup()
```

## counting things

count month/name observations

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  #count the number of observation for each month
  summarize(n_obs = n())
```

we can also use `count()`

```{r}
weather_df %>% 
  count(name, month, name = "n_obs")
```

***NEVER*** use base R's `table`
 - result is not DF
 - not helpful to look at
 
```{r eval = F}
weather_df %>% 
  pull(month) %>% 
  table()
```

other helpful counters

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(),
    #look at the distinct days in each month
    n_days = n_distinct(date))
```

## A digression on 2x2 tables

```{r}
weather_df %>% 
  filter(name!= "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >=5 ~ "not cold",
      TRUE ~ ""
    )
  ) 
  #group_by(name, cold) %>% 
  #summarize(count = n()) 
  
  #quicker way to create 2 by 2 table
  #janitor::tably(name, cold)
```


## General summaries

You can do lots of summaries

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = T),
    mean_prcp = mean(prcp, na.rm = T),
    median_tmin = median(tmin, na.rm = T))
```


This is a df so we can make a plot

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = T),
    mean_prcp = mean(prcp, na.rm = T),
    median_tmin = median(tmin, na.rm = T)
  ) %>% 
  ggplot(aes(x = month, y = mean_tmax)) +
  geom_point() +
  geom_line()
```


Suppose you want to summarize many columns

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(across(prcp:tmin, mean))
```

Reminder: sometimes your results are easier to read in another format

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) %>% 
  # round the table results to 1 digit num
  knitr::kable(digits = 1)
```

## `group_by()` and `mutate()`

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax,na.rm = T),
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```


## what about window functions?

ranking...

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  #rank the minimum of max temp and display the day
  mutate(temp_rank = min_rank(tmax)) %>% 
  #get the coldest day of each month for each location
  filter(temp_rank ==1)
```

lag - shift data down the rows

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(temp_change = tmax - lag(tmax)) %>% 
  summarize(
    #max temp difference for each location & sd
    temp_change_max = max(temp_change, na.rm = T),
    temp_change_sd = sd(temp_change, na.rm = T)
  )
```

























